One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-14 paper/response.txt

I thank the reviewer for a timely report and for providing important 
comments. I have significantly re-structured the 
paper to re-orientate this work in order to address
the issues raised by the reviewer. The revisions are made in bold,
which I further explain below in chronological order. Response to the
reviewer's specific comments follow afterwards.

@@

**********************************************************************

Abstract:

This paper describes a potentially destabilising effect of adopting
a locally isothermal equation of state. The fact that  
linear disturbances do not generally conserve angular momentum in
locally isothermal discs has not been appreciated. I believe it is 
worth pointing out this property of locally isothermal disks given 
that such models are still widely used to model protoplanetary discs.

The basic idea behind the background torque is now presented first. 
The numerical simulations then serve to demonstrate
this destabilisation mechanism to make one-armed spiral patterns 
grow in locally isothermal discs.

@@

Introduction:

In accordance with the reviewer's concerns, I have removed references
to dead zones in this section. I deferred this issue
to the discussions section.

I now simply give a brief description of possible origins of spiral 
structure in circumstellar discs. I highlight previous works on m=1 or
eccentric modes, since the present simulations give one-armed
spirals. I then state that this paper presents a distinct way to grow
one-armed spirals by exploiting the non-conservation of angular
momentum of linear disturbances in locally isothermal disks.

The new introduction prevents the reader from confusing the results of  
this work with proper gravitational instabilities.

@@

The results on variable temperature profiles (Section 5.2) now follows
the reference case (Section 5.1).

Section 7: Discussion

Here I address most of the reviewer's comments.

I reiterate the one-armed spiral instability is not 
self-gravitational. The radially-structured disc models adopted here permits trapped 
one-armed modes because of disc self-gravity, but it does not grow by 
exerting gravitational torques on other parts of the disc (which 
underpins classic gravitational instabilities).

I checked this by performing a new simulation with a lower disc mass 
such that Q>2.5 everywhere in the disc. Then classic
non-axisymmmetric gravitational instabilities should not occur. However,
this model still develops the one-armed spiral because of the forced
temperature gradient, but it grows more slowly.

@@

Section 7.1:

In fact, simulations performed in the early stages of this project did
include the indirect potential (allowing the star to move), but these produced 
similar results. This confirms that the indirect potential is not related 
to the formation of the one-armed spiral patterns in our disc models.

@@

Section 7.2:

The calculations for this section has been moved in
Appendix C. Here, I state the finding that co-rotation and Lindblad
resonances are not important for the one-armed spiral pattern in our
disc models. This supports the notion that the instability is not
gravitational.

Section 7.3:

Here I discuss possible relevance of the results to protoplanetary
discs.

@@

I state that the destabilising mechanism --- the background
torque --- only exists in the presence of a fixed temperature
gradient. This in turn requires the system to return to its initial
temperature profile quickly when perturbed.

I suggest that outer parts of protoplanetary discs, where rapid
cooling may be possible and in which stellar irradiation can be
important, may be a potential site for the background torque to
operate.

For example, in radiation hydrodynamic simulations performed by
Boley (2009) the outer regions of protoplanetary discs (>100AU)
behaves isothermally because of external irradiation. In that work the
irradiation temperature is uniform so the background torque is
absent. However, it is conceivable that external irradiation can set a 
non-uniform disc temperature profile, 
which may enable the background torque.

@@

(The existence of the background torque and gravitational
instability/fragmentation are not necessarily mutually exclusive.)

The other requirement, more specific to trapping one-armed
spirals, is the radial disc structure. Our disc models can trap
one-armed modes because of the rising Q-profile towards the inner disc
which, in turn, is a result of the adopted surface density bump.

A surface density bump may represent mass built-up at some radial
location in the disc. Protoplanetary disc dead zones are an obvious
example. However, as the reviewer points out, dead zones may not
support the background torque because it cools too slowly.

@@

Section 8: Summaryand conclusions

The summary section has been re-ordered to reflect the new paper
structure.

Seciton 8.1:

At the end of this section I remind the reader that the background torque
is a generic feature of locally isothermal discs. It may be relevant whenever 
the disc temperature is set externally. Here, I mention another possible application in 
Be star discs.

@@

************************************************************************

This is one of those papers that I find quite difficult to review.
There's nothing that I can find that is fundamentally wrong with the
paper.  It's easy to read.  The methods and assumptions are clearly
explained.  It's trying to address a topic that is clearly of interest
(how can we explain spiral structure seen in protostellar
discs). However, it is presenting something that I would regard as
largely unrealistic.   I'll explain my reasoning below.

The simulations themselves consider an isothermal or locally
isothermal equation of state only.  Gammie (2001) and other work since
have shown that the evolution of self-gravitating discs depends both
on the stability criterion and on the balance between heating and
cooling.  If the cooling rate is fast, it is likely to fragment,
otherwise it will settle into a quasi-steady, self-gravitating state.  
************************************************************************
The issue of fragmentation due to rapid cooling, as implied by assuming a 
locally isothermal equation of state, is now mentioned in Section 7.3.1.

*************************************************************************

When considering isothermal simulations, the likely evolution is one
of two things.  It's either initially so unstable that the disc
quickly fragments, or it's initially stable enough that any initial
perturbations simply drives the system towards being even more stable
and the instability quickly weakens through rapid mass transfer (see
the work by Laughlin & Bodenheimer 1994/1996).  There is no real way
to drive a long-lived quasi-steady, non-fragmenting, self-gravitating
state because this is only possible if the system can remain
quasi-stable though balancing heating and cooling (see, for example,
Lodato & Rice 2004).  
************************************************************************
Our disc models have radial structure. It is not clear how to  
apply results obtained the above studies that use smooth discs to the 
present work.

In a smooth disc non-axisymmetric gravitational instabilities can 
operate when Q<1.5-1.7. This is satisfied in our reference disc model, 
but this region is rather narrow, about 10 scale heights wide within the 
surface density bump. It is not obvious that the high-m modes 
observed early on in this simulation is a gravitational instability 
(Fig. 3, right panel), because the radial disc structure may permit 
other hydrodynamical shear instabilities (Lovelace et al. 1999).

In any case, the high-m modes saturate at small amplitude (<0.01) and no 
fragmentation is seen, even though the discs are (locally) isothermal. 
Eventually a one-armed spiral grows, but as explained above, this is not 
a gravitational instability in the usual sense.

Furthermore, fragmentation of the one-armed spiral is unlikely. Durisen

### 2015-01-14 paper/model.tex
writing response to referee
writing response

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state {\bf in the perturbed disk}. {\bf  We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}

@@

\subsection{Unstable interaction between low-frequency modes 
  and the background disc due to an imposed temperature gradient}
{\bf 
Here we show that the torque density acting on a low-frequency, 
local mode due to the background temperature gradient can be negative, which 
would enforce the mode, because it has negative angular momentum.  
}

>>>>

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state {\bf in the perturbed disc}. {\bf  We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}

@@

\subsection{Unstable interaction between low-frequency modes 
  and the background disc due to an imposed temperature gradient}
{\bf 
Here we show that the torque density acting on a 
local mode due to the background temperature gradient can be negative, which 
would enforce low-frequency modes, because they have negative angular momentum.  
}
